---
title: "Lab 1, Part 1 — R, RStudio, the Tidyverse and `sf`"
subtitle: "GEOG 413 — Advanced GIS"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-tools: true
    code-copy: true
    embed-resources: false
    theme: default
execute:
  echo: true
  message: false
  warning: false
  cache: true
  fig-width: 8
  fig-height: 5
editor_options:
  chunk_output_type: console
---

```{=html}
<style>
body { font-family: Calibri, "Segoe UI", Arial, sans-serif; }
</style>

options("mapviewBasemaps" = c("Esri.WorldGrayCanvas","OpenStreetMap", "Esri.WorldImagery", "OpenTopoMap"))

```

## Purpose {.unnumbered}

Part 1 gets you from a fresh install to a rendered document that reads spatial data, summarises it and maps it. Part 2 (*Simple Features in R with sf*) then works through the `sf` package properly. Everything runs on `data/chilako.gpkg` from the course site.

Work through it in order, typing the code into your own `part1.qmd`. Render at the end of every section.

## Welcome to RStudio

### RStudio basics

**R** is the language and the engine. **RStudio** is the workbench around it (The "IDE"). The RStudio IDE typically has four panes that can be customized. Here is what they do: 

| Pane | Use it for |
|---|---|
| Source (top-left) | Scripts and `.qmd` files — things you keep |
| Console (bottom-left) | Trying one line; nothing here is saved |
| Environment / History (top-right) | What objects exist right now |
| Files / Plots / Packages / Help (bottom-right) | Browsing, looking at maps, reading docs |

```{r}
#| label: basics
# This is a comment. R ignores everything after '#'.
2 + 2 # put the cursor on this line and press Ctrl+Enter
x <- 10 # assignment; x appears in the Environment pane (similar to '=')
x * 3
```

```{r}
#| label: help
#| eval: false
?mean # help for one function (or highlight a function and press F1)
??"standard deviation" # search all installed help
```

Shortcuts to learn today:

| Keys | Does |
|---|---|
| Ctrl+Enter | run the current line or selection |
| Ctrl+Shift+Enter | run the whole chunk |
| Ctrl+Shift+M | insert a pipe `|>` |
| Alt+- | insert left-assignment `<-` 
| Ctrl+Shift+C | comment / uncomment |
| Ctrl+Shift+K | render the document |
| Ctrl+Shift+F10 | restart R |
| Tab | autocomplete — start typing a function then Tab |

### RStudio settings (no `.RData`!)

*Tools → Global Options → General*:

- Your script should rebuild everything from the data files, not from restoring previously assigned variables. A saved workspace hides the fact that a script may no longer work. Another way to do this is : *Session → Restart R*, then run your script from top to bottom.

  - **Untick** *Restore most recently opened project at startup *
  - **Untick** *Restore .RData into workspace at startup*
  - Set *Save workspace to .RData on exit* to **Never**

- *Tools → Global Options → Code*: 

  - Turn on *Use native pipe operator, |>*. Some of you may be familiar with the `dplyr` pipe (`%>%`), now there is a pipe built into R!
  - Under *Display*, tick *Show line numbers* and turn on *Use rainbow parentheses*!

### The Working Directory

The working directory can bee seen at the top of the Console pane. You can also find it through code with `getwd()`, and you can change it with `setwd()`. Keep in mind file paths in R need to have 
- Wrong: `K:\GEOG413_git\GEOG_413_Lab1`
- Right: `K:/GEOG413_git/GEOG_413_Lab1`
Managing the working directory is important! 

### RStudio Projects

Projects in R Studio, make it so that all code in the project uses a common working directory, which helps make the code more portable. 

*File → New Project → * ... there are three options. 

### Get data for Lab

```{r}
setwd("K:/")
system("git clone https://github.com/bevingtona/GEOG413_git") 
```


```{r}
#| label: project
getwd()
list.files(path = "data")
```

```{r}
#| label: setwd
#| eval: false
# Avoid this! It exists on exactly one computer.
setwd("C:/Users/mycomputer/OneDrive/UNBC/GEOG413/lab")
```

A tidy project:

```
geog413-lab/
├── data/        inputs (add to .gitignore — never commit data to GitHub)
├── outputs/     what you make
├── part1.qmd
└── geog413-lab.Rproj
```

```{r}
#| label: outputs-dir
dir.create("outputs", showWarnings = FALSE)
```

### Managing libraries

```{r}
#| label: install
#| eval: false
install.packages("sf") # once per computer
install.packages(c("dplyr", "ggplot2", "mapview", "units", "ggspatial"))
```

```{r}
#| label: library
library(sf) # once per session
packageVersion("sf")
```

`pkg::fun` calls a function without `library()`, and resolves name clashes (`dplyr::filter` vs `stats::filter`). `sessionInfo()` lists everything loaded — paste it into bug reports. `update.packages(ask = FALSE)` occasionally, not the night before a deadline.

### `.R` vs `.Rmd` vs `.qmd`

**`.R`** — plain code, run top to bottom with `source("analysis.R")`. For helper functions and data prep.

```{r}
#| label: dot-r
#| eval: false
library(sf)
ws <- st_read("data/chilako.gpkg", layer = "watershed")
plot(st_geometry(ws))
```

**`.qmd`** — Quarto: prose and code together, rendered to HTML with Ctrl+Shift+K.

````markdown
---
title: "Chilako watershed"
format: html
---

The watershed boundary comes from the Freshwater Atlas.

```{{r}}
library(sf)
ws <- st_read("data/chilako.gpkg", layer = "watershed")
plot(st_geometry(ws))
```

It covers `r knitr::inline_expr("round(ws$AREA_HA)")` hectares.
````

The last line is *inline code*: a number in your prose that recalculates when the data change.

**`.Rmd`** is the older R Markdown. Same idea, same chunks. The visible differences: chunk options go inside the chunk as `#| label: read-ws` in Quarto rather than `{r read-ws, echo=FALSE}` in R Markdown; Quarto also renders Python and Julia and has better defaults for HTML, PDF, slides and websites. **Write `.qmd`.** Read `.Rmd` when you find it.

## Basic R

### Data manipulation

**Vectors** hold many values of one type.

```{r}
#| label: vectors
years  <- c(2015, 2018, 2021)              # numeric
sites  <- c("CHK-01", "CHK-02", "CHK-03")  # character
active <- c(TRUE, FALSE, TRUE)             # logical

years[2]                                   # index from 1, not 0 like python
years[years > 2016]                        # logical subsetting
length(years)           
class(years)          
```

A **data.frame** is a table: each column is a vector of one type.

```{r}
#| label: dataframe
loggers <- data.frame(site = sites, year = years, active = active)
loggers
loggers$year                               # one column, as a vector
loggers[2, ]                               # one row
loggers[loggers$active, "site"]            # rows where active, column site
nrow(loggers); ncol(loggers); names(loggers)
str(loggers)
```

A **list** holds anything in any mix. Geometry columns in `sf` are lists — you'll meet them in Part 2.

```{r}
#| label: list
config <- list(watershed = "Chilako River", crs = 3005, years = years)
config$watershed
config[["crs"]]
str(config)
```

**Types and conversion.**

```{r}
#| label: types
x <- "2018"
class(x)
as.numeric(x) + 1
as.character(2018)
as.numeric("abc")                          # NA with a warning: check your data
```

**Dates** are strings until you say otherwise.

YYYY-MM-DD PREFERRED!!!!! 
%Y-%m-%d

```{r}
#| label: dates
d <- "2018-07-15"
class(d)
d <- as.Date(d)                            # ISO format needs no template
class(d)
d + 30
format(d, "%Y")                            # the year, as text
as.numeric(format(d, "%Y"))                # ...as a number
as.Date("15/07/2018", format = "%d/%m/%Y") # non-ISO: describe the pieces
```

**Factors** are categories with fixed levels; ggplot uses them for ordering.

```{r}
#| label: factors
cause <- factor(c("Lightning", "Person", "Lightning", "Lightning", "Lightning", "Lightning", "Lightning"))
print(cause)
levels(cause); 
table(cause)
```

**Missing values** poison arithmetic unless you say what to do with them.

```{r}
#| label: na
v <- c(3, NA, 7)
mean(v)
mean(v, na.rm = TRUE)
```

### Create a table

```{r}
#| label: tibble
library(dplyr)
loggers <- tibble( # a tibble is a data.frame that prints better
  site   = c("CHK-01", "CHK-02", "CHK-03"),
  elev_m = c(720, 860, 1010),
  lon    = c(-123.30, -123.65, -123.95),
  lat    = c(53.75, 53.55, 53.40))
loggers
```

Write it out and read it back:

```{r}
#| label: csv
write.csv(loggers, "data/loggers.csv", row.names = FALSE)

# Base R version
loggers <- read.csv("data/loggers.csv")
loggers
str(loggers)
summary(loggers)

# Tidy R version
loggers <- readr::read_csv("data/loggers.csv")
loggers
str(loggers)
summary(loggers)
```

`View(loggers)` opens a spreadsheet view in RStudio. In a rendered document, use a proper table:

```{r}
#| label: kable
# Table for rendering into a document:
knitr::kable(loggers, digits = 2, caption = "Stream-temperature loggers")
```

### Basic plotting

Base R plotting needs no packages and is fine for looking. Not for publishing.

```{r}
#| label: base-plot
plot(x = loggers$elev_m, 
     y = loggers$lat, 
     pch = 16, 
     col = "steelblue",
     xlab = "Elevation (m)", 
     ylab = "Latitude", 
     main = "Loggers")
text(loggers$elev_m, loggers$lat, loggers$site, pos = 1)
```

```{r}
#| label: base-plot-2
#| fig-height: 3.5
par(mfrow = c(1, 3))
hist(rnorm(100))
barplot(c(a = 3, b = 7, c = 5))
boxplot(elev_m ~ site, data = loggers)
```

```{r}
#| label: base-save
#| eval: false
png("outputs/loggers_base.png", width = 800, height = 600)
plot(loggers$elev_m, loggers$lat, pch = 16)
dev.off()
```

### Tidyverse syntax

We'll use the fire attribute table. For now the geometry is dropped; it comes back in the `sf` section.

```{r}
#| label: fires-table
fires <- st_read("data/chilako.gpkg", layer = "fires", quiet = TRUE) |> st_drop_geometry()
head(fires)
```

**The pipe** `|>` reads as "take this, *then* do that". Ctrl+Shift+M types it.

```{r}
#| label: pipe
# Without: read from the inside out
head(arrange(filter(fires, FIRE_YEAR >= 2000), FIRE_YEAR))

# With: read top to bottom
fires |>
  filter(FIRE_YEAR >= 2000) |>
  arrange(FIRE_YEAR) |>
  head()
```

**`select()`** — columns.

```{r}
#| label: select
fires |> select(FIRE_NUMBER, FIRE_YEAR) |> head(3)
fires |> select(-FIRE_CAUSE) |> head(3)
fires |> select(starts_with("FIRE_Y")) |> head(3)
```

**`filter()`** — rows.

```{r}
#| label: filter
fires |> filter(FIRE_YEAR >= 2010) |> nrow()
fires |> filter(FIRE_YEAR >= 2010, FIRE_CAUSE == "Lightning") |> nrow()   # AND
fires |> filter(FIRE_YEAR < 1950 | FIRE_YEAR > 2015) |> nrow()            # OR
fires |> filter(!is.na(FIRE_CAUSE)) |> nrow()
```

**`mutate()`** — new or changed columns.

```{r}
#| label: mutate
fires |>
  mutate(decade = floor(FIRE_YEAR / 10) * 10,
         recent = FIRE_YEAR >= 2010) |>
  head()
```

**`group_by()` + `summarise()`** — one row per group.

```{r}
#| label: summarise
fires |>
  group_by(FIRE_CAUSE) |>
  summarise(n = n(),
            first = min(FIRE_YEAR),
            last  = max(FIRE_YEAR))
```

All together, with `arrange()`:

```{r}
#| label: together
fires |>
  mutate(decade = floor(FIRE_YEAR / 10) * 10) |>
  group_by(decade, FIRE_CAUSE) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  head()
```

`count()` is `group_by()` + `summarise(n = n())` in one call:

```{r}
#| label: count
fires |> count(FIRE_CAUSE, sort = TRUE)
```

### Tidyverse and ggplot2

The grammar: **data** + **aes** (which columns map to what) + **geom** (how to draw it).

```{r}
#| label: cutblocks-table
library(ggplot2)
cutblocks <- st_read("data/chilako.gpkg", layer = "cutblocks", quiet = TRUE) |>
  mutate(area_ha = as.numeric(units::set_units(st_area(geom), ha))) |>
  st_drop_geometry()

ggplot(cutblocks, aes(x = HARVEST_YEAR)) +
  geom_histogram(binwidth = 1)
```

Pipe straight from a summary into a plot:

```{r}
#| label: gg-col
cutblocks |>
  group_by(HARVEST_YEAR) |>
  summarise(area_ha = sum(area_ha)) |>
  ggplot(aes(HARVEST_YEAR, area_ha)) +
  geom_col() +
  labs(x = "Harvest year", y = "Area (ha)", title = "Harvest history, Chilako River") +
  theme_minimal()
```

Boxplot by group, on a log axis:

```{r}
#| label: gg-box
fires <- st_read("data/chilako.gpkg", layer = "fires", quiet = TRUE) |>
  mutate(area_ha = as.numeric(units::set_units(st_area(geom), ha))) |>
  st_drop_geometry()

ggplot(fires, aes(FIRE_CAUSE, area_ha)) +
  geom_boxplot() +
  scale_y_log10()
```

Scatter with a smoother:

```{r}
#| label: gg-scatter
ggplot(cutblocks, aes(HARVEST_YEAR, area_ha)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  scale_y_log10()
```

Lines, one per group:

```{r}
#| label: gg-line
fires |>
  mutate(decade = floor(FIRE_YEAR / 10) * 10) |>
  count(decade, FIRE_CAUSE) |>
  ggplot(aes(decade, n, colour = FIRE_CAUSE)) +
  geom_line() + geom_point()
```

Facets, one panel per group:

```{r}
#| label: gg-facet
ggplot(fires, aes(area_ha)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ FIRE_CAUSE, scales = "free_y") +
  scale_x_log10()
```

Save:

```{r}
#| label: gg-save
p <- ggplot(cutblocks, aes(HARVEST_YEAR)) + geom_histogram(binwidth = 1)
ggsave("outputs/harvest_hist.png", p, width = 8, height = 5, dpi = 200)
```

## Introduction to `sf`

### Read/write

```{r}
#| label: sf-read
st_layers("data/chilako.gpkg")

ws        <- st_read("data/chilako.gpkg", layer = "watershed", quiet = TRUE)
streams   <- st_read("data/chilako.gpkg", layer = "streams",   quiet = TRUE)
lakes     <- st_read("data/chilako.gpkg", layer = "lakes",     quiet = TRUE)
cutblocks <- st_read("data/chilako.gpkg", layer = "cutblocks", quiet = TRUE)
fires     <- st_read("data/chilako.gpkg", layer = "fires",     quiet = TRUE)

ws
```

Read the header before anything else: number of features and fields, geometry type, bounding box, CRS.

```{r}
#| label: sf-write
st_write(ws, "outputs/ws.gpkg", layer = "watershed", delete_layer = TRUE, quiet = TRUE)
```

```{r}
#| label: sf-write-other
#| eval: false
st_write(ws, "outputs/ws.shp")        # can, but don't: 10-character names, 3+ files, lossy CRS
st_write(st_transform(ws, 4326), "outputs/ws.geojson")   # web formats want WGS 84
```

### Geometry types

```{r}
#| label: geom-types
st_geometry_type(ws)
st_geometry_type(streams) |> unique()
```

Each type can be built by hand:

```{r}
#| label: geom-build
pt   <- st_point(c(1150000, 980000))
ln   <- st_linestring(rbind(c(1120000, 970000), c(1180000, 1010000)))
poly <- st_polygon(list(rbind(c(0, 0), c(1, 0), c(1, 1), c(0, 1), c(0, 0))))   
class(pt)

# add a CRS and attributes
line_sf <- st_sf(name = "transect", geometry = st_sfc(ln, crs = 3005)) 

line_sf
```

Cast between types; make points from polygons:

```{r}
#| label: geom-cast
nrow(streams)
streams |> st_cast("LINESTRING") |> nrow()      # multi → single: more rows
st_centroid(lakes) |> head(3)
```

### Predicates and operations

```{r}
#| label: predicates
#| cache: false

library(mapview)
big_fire <- fires |> slice_max(st_area(geom), n = 1)

st_intersects(big_fire, cutblocks)              # sparse list of matching indices
st_intersects(big_fire, cutblocks, sparse = FALSE) |> sum()

cutblocks[big_fire, ] |> nrow()                 # subset: intersects by default
cutblocks[big_fire, , op = st_within] |> nrow() # any predicate via op
cutblocks |> st_filter(big_fire) |> nrow()      # same, pipeable
cutblocks_bigfire <- cutblocks |> st_filter(big_fire)

mapview(big_fire, col.regions = "red") + 
  mapview::mapview(cutblocks_bigfire, col.regions = "green")

cutblocks_bigfire_nearlake <- cutblocks_bigfire |> 
  st_is_within_distance(lakes, units::set_units(200, m))

mapview(big_fire, col.regions = "red") + 
  mapview(cutblocks_bigfire, col.regions = "green") + 
  mapview(cutblocks_bigfire[lengths(cutblocks_bigfire_nearlake) > 0, ], col.regions = "blue")
```

Measures come back with units:

```{r}
#| label: measures
st_area(ws) |> units::set_units(km^2)
st_length(streams) |> sum() |> units::set_units(km)
```

Geometry operations:

```{r}
#| label: geom-ops
#| eval: false
st_union(cutblocks)                             # dissolve everything to one
st_intersection(cutblocks, big_fire)            # clip
st_difference(big_fire, st_union(cutblocks))    # erase
st_buffer(lakes, 200)
st_make_valid(cutblocks)
```

### Tidyverse

The dplyr verbs work on `sf` objects and the geometry comes along.

```{r}
#| label: sf-tidy
cutblocks |> filter(HARVEST_YEAR >= 2010) |> nrow()
cutblocks |> select(HARVEST_YEAR) |> head(2)
cutblocks |> mutate(area_ha = st_area(geom)) |> head(2)
```

`summarise()` on an `sf` object unions the geometry of each group — a dissolve:

```{r}
#| label: sf-dissolve
cb_decade <- cutblocks |>
  mutate(decade = floor(HARVEST_YEAR / 10) * 10) |>
  group_by(decade) |>
  summarise(n = n())
cb_decade
```

Drop the geometry when you only want a table:

```{r}
#| label: sf-drop
cutblocks |> st_drop_geometry() |> count(HARVEST_YEAR) |> head(3)
```

A CSV with coordinates becomes spatial with `st_as_sf()` — say which columns, and what CRS they're in:

```{r}
#| label: csv-to-sf
loggers_sf <- read.csv("data/loggers.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)
loggers_sf
```

### Plotting with ggplot2

```{r}
#| label: gg-map
#| fig-height: 8
ggplot() +
  geom_sf(data = ws, fill = "grey95") +
  
  geom_sf(data = fires, aes(fill = "Fire"), alpha = 0.4, colour = NA) +
  scale_fill_manual(name = "Wildfire", values = "red") +
  ggnewscale::new_scale_fill() +
  
  geom_sf(data = cutblocks, aes(fill = HARVEST_YEAR), colour = NA) +
  scale_fill_viridis_c(name = "Harvest year") +
  ggnewscale::new_scale_fill() +
  
  geom_sf(data = streams, aes(linewidth = STREAM_ORDER), colour = "steelblue") +
  scale_linewidth(name = "Stream Order", range = c(0.2, 1)) +
  
  geom_sf(data = lakes, fill = "steelblue", colour = NA) +
  
  # ADD LOGGER POINTS AND LABELS 
  geom_sf(data = loggers_sf, colour = "red", size = 2) +
  geom_sf_label(data = loggers_sf, aes(label = site), nudge_y = 2000, size = 3) +
  
  # ADD STYLE
  ggspatial::annotation_scale(location = "bl") +
  ggspatial::annotation_north_arrow(location = "tr") +
  labs(title = "Chilako River watershed", x = "", y = "") + 
  theme_bw()

ggsave("outputs/chilako_map.png", width = 10, height = 8, dpi = 300, bg = "white")
```

### Plotting with mapview

For exploring, not for reports.

```{r}
#| label: mapview
#| cache: false
library(mapview)
mapview(cutblocks, zcol = "HARVEST_YEAR")
```

```{r}
#| label: mapview-layers
#| cache: false
mapview(ws, alpha.regions = 0) +
  mapview(fires, col.regions = "orange") +
  mapview(lakes, col.regions = "steelblue") +
  mapview(loggers_sf, col.regions = "red")
```

Two things to know: mapview reprojects to Web Mercator for display — your data are still in 3005 — and its output cannot be cached, so any chunk that calls it needs `#| cache: false` if you turn caching on.

## Before you leave {.unnumbered}

- `part1.qmd` renders from a fresh session (Ctrl+Shift+F10, then Ctrl+Shift+K)
- `outputs/` has `harvest_hist.png` and `chilako_map.png`
- Committed and pushed; `data/` is not in the repo

Next: Part 2, *Simple Features in R with sf*.
